The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-22 Independence_CH/lessons.tex
Slight revision of Lessons

A similar experience, but on an opposite side of the formalization
spectrum happened to the Liquid Tensor Experiment as described by Scholze
in \cite{LTE2021}. People involved in the formalization simply pushed
their way to reach the summit, formalizing lemma after lemma. They
actually wrote the blueprint for that formalization \emph{afterwards}
it was complete! From time to time, we were also frenziedly trying to
get the results formalized, going beyond what we had planned.

@@

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families;
  thus limiting us to the plain ccc, and avoiding the relativization
  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $(<\omega+1)$-closed); and
\end{itemize}
In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isatt{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

@@

It is noteworthy that although the “math” of the construction of a
model of $\ZFC+\neg\CH$ was already in place by the end of November
2020, it was only 9 months later that we were able to finish the
formalization of that result. The missing pieces were essentially
bureaucracy. Some of the material filed under this category comprises:
\begin{itemize}
\item permutation of indices and calculation of arities of
  internalized formulas;
\item proving that certain constructions belong to the relevant
  models;
\item (required for the above) showing that particular instances of
  separation and replacement hold in the ground model;
\item etc.
\end{itemize}

Some of those proofs were almost copy-pasted once and again with minor
variants; this would usually be relegated to some function in the
meta-language, but we were unable to do this due to our limitations in
programming Isabelle/ML\footnote{%
  On the other hand, our inability to automate proofs of replacement
  instances paved the way for identifying which were the ones needed for
  forcing! ($\leftarrow$\textbf{put somewhere, don't know if this is the right
    place}).}.

>>>>

A similar experience, but on an opposite side of the formalization
spectrum happened to the Liquid Tensor Experiment \cite{nature-lte} as described by Scholze
in \cite{LTE2021}. People involved in the formalization simply pushed
their way to reach the summit, formalizing lemma after lemma. They
actually wrote the blueprint for that formalization \emph{afterwards}
it was complete! From time to time, we were also frenziedly trying to
get the results formalized, going beyond what we had planned.

@@

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families;
  thus limiting us to $\kappa=\aleph_1$ case of the $\kappa$-chain condition, and dodging the relativization
  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $(<\omega+1)$-closed, that is, under plain $\omega$-sequences).
\end{itemize}
In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isatt{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

@@

It is noteworthy that although the “math” of the construction of a
model of $\ZFC+\neg\CH$ was already in place by the end of November
2020, it was only 9 months later that we were able to finish the
formalization of that result. The missing pieces were essentially
bureaucracy. Some of the material filed under this category comprises:
\begin{itemize}
\item permutation of indices and calculation of arities of
  internalized formulas;
\item proving that certain constructions belong to the relevant
  models;
\item (required for the above) showing that particular instances of
  separation and replacement hold in the ground model.
\end{itemize}

Some of those proofs were almost copy-pasted once and again with minor
variants; this would usually be relegated to some function in the
meta-language, but we were unable to do this due to our limitations in
programming Isabelle/ML\footnote{%
  On the other hand, our inability to automate proofs of replacement
  instances paved the way for identifying which were the ones needed for
  forcing!}.
